---
title: "Movie star names and baby names"
subtitle: "SI 618 Winter 2017 Project Report Part-b"
author: "Chong Li"
date: "4/18/2017"
output: pdf_document
---

# Background and Motivations

In China, there is generally no pool of names for babies. However, in western countries, parents often name their babies from a pool of names. I am curious about how other factors influence the baby names. I am also surprised by American's enthusiasm for movies. So I decide to see how movie star names could influence the names of babies. I would like to see how star names in IMDB top movies and academy award winning star names influence the baby's name in the corresponding years.

In the first part of my projects, I have found that there seems to be a significant relationship between babies named and movie stars. Therefore, I plan to explore further the relationship between the two in terms of **gender**, **movie genre**, and **year**, and find out what factors is most influential in naming babies. For example, is naming babies after movie stars becoming more popular or less popular? What movie genre has the most significant influence? Do female movie star names have an influence on boy babies?

The four questions I choose to explore are as follows:

1. **What are the influential first names?** Specifically, I will explore the most influential names and how they influence boy babies or girl babies.

2. **Is there cross-gender influence?** I want to know if male movies stars had great influence on girl names and vice versa. 

3. **What types of movies are more influential?** I also want to know if a specific movie genre has more influence than others.

4. **Is naming after movie stars more popular or less?** I will explore the trend of naming after movie stars is becoming more popular across the years or less. In particular, in the peak years when movie star names were most influential, what movie genres were more influential?

# Data Source

For the actors/actresses' names, I used the IMDB 5000 movies dataset, which is available in CSV format on [kaggle imdb 5000 movies](https://www.kaggle.com/deepmatrix/imdb-5000-movie-dataset). This dataset contains three leading actors/actresses' names of each movie, the publishing year, movie genre, director, rating counts, ect. 

For the baby names, I found the dataset in CSV at [kaggle us baby names](https://www.kaggle.com/kaggle/us-baby-names). This dataset contains the ranking of baby names from 1880 to 2016. In each year, I can get the ranking and number of names for both boys and girls. In this dataset, I can even further explore name trends by state.

In addition, I retrived the academy award best leading actors' and best leading actresses' names from 1927 to 2016 by parsing the Wikipedia page using urllib and beautiful soup.

The important variables across the three datasets are name(string), year(string), movie genre(string), and gender(string).

# Methods

## (a) How did you manipulate the data to prepare it for analysis?  

1. **Academy award dataset**: I retrived this data set from wikipedia page and stored names of actors and actresses in two dictionaries. The keys are the years and the values are names. I used this format so that I could easily retrieve dat by year and compare the trend across years combining different datasets.

2. **Baby name dataset**: The dataset I got is in csv format with a (name, number) pair on each line. I manipulated the data into two dictionaries (boy and girl). The keys are also year and the values are a list of tuples (year, number that year).

3. **IMDB dataset**: I cleaned this dataset and only retrived the top 10 movie star names per year by ranking. I retrieved two dictionaries from this dataset: name dictionary and genre dictionary. Both dictionaries have year as the key.

4. **Combining dataset**. I combined the two datasets and got a new dictionary with year as the key and a list of names as the value. Then I counted the occurence of different first names in the values and retrived a new data set of first names that appeared 5 or more times in the combined dictionary.

5. **Baby name numbers**. Then I retrieved the corresponding baby name numbers across 1927 to 2015 for the 5+ occurences names mentioned above in csv files. The format is (year, female name number, male name number).

For the first question, I used the **Combining dataset** to show the rank of occurences of different names. I also used the **Baby name numbers** and **Combining dataset** to calculate the **influence rate dataset** (specified in (d)). I ranked the **influence rate dataset** and got the most influential names.

For the second question, I continued to use the **influence rate dataset** but filtered and manipulated the data across genders.

For the third question, I distilled a **names' genre and influnece rate dataset** in the format of CSV ("genre_pairs.csv") and performed analysis between movie genre and influences.

For the last question, I used the **Baby name numbers** and manipulated the data to be easily sorted by year so that I can calulate the tread across years.

## (b) How did you handle missing, incomplete, or noisy data?

The datasets are generally clean and I do not need to implement any data cleaning methods. There is also no missing data observed across the datasets. Therefore, I did not perform much data handling across the four questions.

Specifically, for the first question "what are the influential names", I generated a random int between 50 to 100 when the occurence of a certian name dropped to be less than 100. The reason is that by doing so, the influence will not fluctuate too much due to small sample size (I specified how to calculate the influence in (d)).

## (c) How did you perform data analysis in code, i.e. briefly describe the workflow of your source code

The first step is to prepare distill ready-to-use datasets from the three raw datasets. I manipulated the source datasets in python and returned different CSV files for further analysis in R. These CSV files include files containing baby name trends, movie star names, movie genres, ect.

Then I moved into R and utilized ggplot2 libraries to draw different plots such as histrograms, bar charts, boxplots, line plots, and jitter plots. I also filtered the datasets whenever necessary. 

As for the sequence of data analysis, I first analyzed more general questions such as the most influential movie star names and moved to more specific questions such as cross-gender influence, movie genre influence. Finally, I analyzed the overal trend of movie star names' influence by using jitter plots and fitting a linear regression model.

## (d) What challenges did you encounter and how did you solve them?

As my four questions are highly related, the challenges I encounted for each questions are very similar. The challenges are as follows:

#### Dealing with instances happening in recent years

As I am analyzing the datasets, I found that some data are very new, e.g. from 2014 to 2016. When dealing with these data, it is impossible to find if the name has an influence since I cannot find the number of babies who have the same name. To solve this, I discarded any name after year 2013 since the baby names dataset covers until 2016.

#### Small data

For the baby name datasets, the number of babies with a certain name can be very small in some years, which will result in big increase/drop rate. To solve this problem, I gave the names with only a small number a indicator of "-1" and ignored them when doing data analysis. In this way, I was basically dealing with relatively large numbers and ruled out extreme cases.

#### Deciding the influence rate

In my analysis, it is very important to define how influential a movie star name is. With only the year when a movei star is successful and the number of baby names every year, it is hard to calculate the influence rate of a movie star name. After further exploration and consideration, I defined the influence rate as the sum of increase percentage of three years after the name appeared minus the sum of the increase percentage of three years before the name appeared. When a name is less popular, the increase percentage will be negative (Please refer to my code "part3.py" for the specific calculation methods). In this way, I am considering the trend of a name in a 6-year period and how a movie star with the same name influenced the new-born babies' names.


# Analysis and Results

## 1. Who are the influential first names?

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(gplots)
library(data.table)
library(cluster)
library(ggplot2)
```

My plot below shows that the most common influetial movie star first name is 'Robert' for male and 'Jennifer' for female. This is just the occurences of the names instead of their influences.

```{r echo=FALSE}
names_5more<- read.table("5+star_names.csv",head=TRUE,sep = ",", fill = TRUE, comment.char = '\"', quote = '\"', encoding = "UTF-8")
ggplot(data=names_5more, aes(x=reorder(name,-occurences), y=occurences)) + theme(axis.text.x = element_text(angle = 60, hjust = 1)) + geom_bar(stat='identity')+labs(y='Occurences',x='Star Names',title='Occurences of movie star names in the distilled dataset')
```


Based on the results above, I plotted the trend of male babies named 'Robert' across the years in red line and plotted the years when a movie star named 'Robert' was successful. The plot shows that movie star 'Robert' did have positive influence on baby 'Robert's, especailly after 1950s when 'Robert' has a trend of declining, every time movie star 'Robert' appeared, the trend of baby 'Robert' stayed for a few years.

```{r echo=FALSE}
Robert<- read.table("namedata/Robert.csv",head=TRUE,sep = ",", fill = TRUE, comment.char = '\"', quote = '\"', encoding = "UTF-8")
Jack<- read.table("namedata/Jack.csv",head=TRUE,sep = ",", fill = TRUE, comment.char = '\"', quote = '\"', encoding = "UTF-8")
Jennifer<- read.table("namedata/Jennifer.csv",head=TRUE,sep = ",", fill = TRUE, comment.char = '\"', quote = '\"', encoding = "UTF-8")
Jane<- read.table("namedata/Jane.csv",head=TRUE,sep = ",", fill = TRUE, comment.char = '\"', quote = '\"', encoding = "UTF-8")

ggplot(data=Robert,aes(x=Robert,y=male_count))+geom_line(color="red")+geom_vline(xintercept=c(1951, 1939, 1983, 1983, 1980, 1988, 1966, 1966, 1967, 1975, 1974, 1974, 1973, 1972, 2005),alpha=I(3/5)) + labs(x="Year", y="Number of boy babies named Robert")
##ggplot(data=Jack,aes(x=Jack,y=male_count))+geom_line(color="red")+geom_vline(xintercept=c(1957, 1958, 1989, 1997, 1968, 1962, 1975, 1973, 1970, 2006, 2007),alpha=I(3/5))
```

I also plotted the relationship in 'Jennifer'. As a result, there is no clear correlation between movie star 'Jennifer' and baby 'Jennifer's.

```{r echo=FALSE}
ggplot(data=Jennifer,aes(x=Jennifer,y=male_count))+geom_line(color="red")+geom_vline(xintercept = c(1943, 1946, 2012, 1984, 2001)) + labs(x="Year", y="Number of girl babies named Jennifer")
##ggplot(data=Jane,aes(x=Jane,y=male_count))+geom_line(color="red")+geom_vline(xintercept = c(1948, 1978, 1971))
```


Then I ranked the movie star names by their influence rate (specified in methods (d)). As shown below, the three most influence names for boy babies are Errol, Dana, Gary, Dirk, and Leaonardo.

```{r echo=FALSE}
influence_rate<- read.table("influence_rate.csv",head=TRUE,sep = ",", fill = TRUE, comment.char = '\"', quote = '\"', encoding = "UTF-8")

influence_rate<-as.data.table(influence_rate)

influence_rate<-influence_rate[influence_rate$Year<2014]

influence_rate_top20m <- influence_rate[order(-InfluenceToBoy)][1:20,]
influence_rate_top20f <- influence_rate[order(-InfluenceToGirl)][1:20,]

ggplot(data=influence_rate_top20m, aes(x=reorder(FullName,-InfluenceToBoy), y=InfluenceToBoy,fill=FirstName)) + theme(axis.text.x = element_text(angle = 60, hjust = 1)) + geom_bar(stat='identity',alpha=I(4/5))+labs(y='Influence Rate',x='Star Names',title='Top 20 boy influence rate of male movie stars')
```

And the five most influential names for girl babies are Jessica, Ryan, Geraldine, Simone, and Jennifer.

```{r echo=FALSE}
ggplot(data=influence_rate_top20f, aes(x=reorder(FullName,-InfluenceToGirl), y=InfluenceToGirl,fill=FirstName)) + theme(axis.text.x = element_text(angle = 60, hjust = 1)) + geom_bar(stat='identity',alpha=I(4/5))+labs(y='Influence Rate',x='Star Names',title='Top 20 girl influence rate of female movie stars')
```

I also plotted the influence rate intervals for male and female movie star names. From the plots below, I inferred that girl baby names are more influenced by movie stars since the higher influence rates are all credited to girls.

```{r echo=FALSE}
influence_rate_clean <- influence_rate[influence_rate$InfluenceToBoy<=5 & influence_rate$InfluenceToBoy>=-1 & influence_rate$InfluenceToGirl<=5 & influence_rate$InfluenceToGirl>=-1]

ggplot(data=influence_rate_clean,aes(x=InfluenceToBoy)) +geom_histogram(binwidth=0.01,alpha=I(4/5)) + labs(y='Count',x='Infuence Rate',title='Histogram of Influence Rate to Boy names')
```

```{r echo=FALSE}
ggplot(data=influence_rate_clean,aes(x=InfluenceToGirl)) +geom_histogram(binwidth=0.01,alpha=I(4/5)) + labs(y='Count',x='Infuence Rate',title='Histogram of Influence Rate to Girl names')

influence_rate_positive_boy <- influence_rate[influence_rate$InfluenceToBoy>0]
influence_rate_positive_girl <- influence_rate[influence_rate$InfluenceToGirl>0]
```

```{r echo=FALSE}
ggplot(data=influence_rate_positive_boy,aes(x=InfluenceToBoy)) +geom_histogram(binwidth=0.01,alpha=I(4/5)) + labs(y='Count',x='Infuence Rate',title='Histogram of Positive Influence Rate to Boy names')
```

```{r echo=FALSE}
ggplot(data=influence_rate_positive_girl,aes(x=InfluenceToGirl)) +geom_histogram(binwidth=0.01,alpha=I(4/5)) + labs(y='Count',x='Infuence Rate',title='Histogram of Positive Influence Rate to Girl names')
```


## 2. Is there cross-gender influence?

I plotted the top 20 influetial movie stars for boy baby names and girl baby names as shown below. It turns out that only one out of twenty boy influencers is female and the first name of that female star is 'Jean', which is a common boy first name in the U.S. As for girls, four out of twenty influncers are male and all the four names are considered to be boy first names in general. 

Therefore, I inferred that girl names are indeed more influenced by movie stars. In addition, girl names tend to be influenced by male movie stars, but boy names are generally not heavily influenced by female movie stars.

```{r echo=FALSE}
influence_rate_positive_boy_top20 <- influence_rate_positive_boy[order(-InfluenceToBoy)][1:20,]
influence_rate_positive_girl_top20 <- influence_rate_positive_girl[order(-InfluenceToGirl)][1:20,]

influence_rate_positive_boy_top20$Gender <- c('male')
influence_rate_positive_boy_top20$Gender[8] <- 'female'

influence_rate_positive_girl_top20$Gender <- c('female')
influence_rate_positive_girl_top20$Gender[2] <- 'male'
influence_rate_positive_girl_top20$Gender[10] <- 'male'
influence_rate_positive_girl_top20$Gender[13] <- 'male'
influence_rate_positive_girl_top20$Gender[20] <- 'male'
```

```{r echo=FALSE}
ggplot(data=influence_rate_positive_boy_top20,aes(y=InfluenceToBoy,x=reorder(FullName,-InfluenceToBoy),fill=Gender)) +geom_bar(stat='identity',alpha=I(4/5)) + theme(axis.text.x = element_text(angle = 60, hjust = 1)) + labs(x='Name',title="Top influences to boy names by gender")
```

```{r echo=FALSE}
ggplot(data=influence_rate_positive_girl_top20,aes(y=InfluenceToGirl,x=reorder(FullName,-InfluenceToGirl),fill=Gender)) +geom_bar(stat='identity',alpha=I(4/5)) + theme(axis.text.x = element_text(angle = 60, hjust = 1)) + labs(x='Name',title="Top influences to girl names by gender")
```

## 3. What types of movies are more influential?

To analyze the influences of different movie genres, I firstly plotted the number of movie star names in each genre as shown below.

```{r echo=FALSE}
genre_pairs<- read.table("genre_pairs.csv",head=TRUE,sep = ",", fill = TRUE, comment.char = '\"', quote = '\"', encoding = "UTF-8")
```

```{r echo=FALSE}
ggplot(data=genre_pairs,aes(x=reorder(Genre,-table(Genre)[Genre])))+geom_bar(stat="count")+theme(axis.text.x = element_text(angle = 60, hjust = 1)) +labs(x='Genre',y="Count",title="Movie star counts in each genre")
```

Then I used the influence rate calculated earlier for each movie star name and grouped the movie star name by genre. After that, I calculated the average influence rate of each genre and drew two boxplots for both boy names and girl names. From the graphs below, the most influential movie genre for boy names is 'Sci-Fi', followed by 'Action' and 'Adventure'. The most influential genre for girl names is 'Music', followed by 'Fantasy' and 'History'.

```{r echo=FALSE}
genre_pairs <- as.data.table(genre_pairs)
genre_pairs_positive_m<- genre_pairs[genre_pairs$InfluenceToBoy>0]
genre_pairs_positive_f<- genre_pairs[genre_pairs$InfluenceToGirl>0]
```

```{r echo=FALSE}
ggplot(data=genre_pairs_positive_m,aes(x=Genre,y=InfluenceToBoy,col=Genre))+geom_boxplot(outlier.shape=NA) + theme(axis.text.x = element_text(angle = 60, hjust = 1))+labs(title="Genre influences to boy names")

##ggplot(data=genre_pairs_positive_m,aes(x=Genre,y=InfluenceToBoy,fill=Genre))+stat_summary(fun.y="mean", geom="bar") + theme(axis.text.x = element_text(angle = 60, hjust = 1))
##ggplot(data=genre_pairs_positive_m,aes(x=Genre,y=InfluenceToBoy,col=Genre))+geom_jitter() + theme(axis.text.x = element_text(angle = 60, hjust = 1))
```

```{r echo=FALSE}
ggplot(data=genre_pairs_positive_f,aes(x=Genre,y=InfluenceToGirl,col=Genre))+geom_boxplot(outlier.shape=NA) + theme(axis.text.x = element_text(angle = 60, hjust = 1))+labs(title="Genre influences to girl names")
##ggplot(data=genre_pairs_positive_f,aes(x=Genre,y=InfluenceToGirl,fill=Genre))+stat_summary(fun.y="mean", geom="bar") + theme(axis.text.x = element_text(angle = 60, hjust = 1))
```

## 4. Is naming after movie stars more popular or less?

I plotted two scatterplots of the influence rate across the years of all the names to see the trend. As a result, the names with high influence rate decreased over the years for both boys and girls. After fitting a linear regression model, it is shown that movie stars influence to boy names stay unchanged. However, the influence of movie stars to girl names decreased approximately from 0.28 to 0.20.

To conclude, naming after movie stars is becoming less popular, especially for girls.

```{r echo=FALSE}
##ggplot(data=influence_rate_clean,aes(x=Year,y=InfluenceToBoy)) +geom_jitter(aes(group=Year),alpha=I(2/5)) + labs(y='Count',x='Infuence Rate',title='Boy names')
##ggplot(data=influence_rate_clean,aes(x=Year,y=InfluenceToGirl)) +geom_jitter(aes(group=Year),alpha=I(2/5)) + labs(y='Count',x='Infuence Rate',title='Girl names')
```

```{r echo=FALSE}
ggplot(data=influence_rate_positive_boy,aes(x=Year,y=InfluenceToBoy)) +geom_point(alpha=I(2/5)) + labs(x='Year',y='Infuence Rate',title='Boy names Trend')+geom_smooth(method=lm)
```

```{r echo=FALSE}
ggplot(data=influence_rate_positive_girl,aes(x=Year,y=InfluenceToGirl)) +geom_point(alpha=I(2/5)) + labs(x='Year',y='Infuence Rate',title='Girl names Trend')+geom_smooth(method=lm)
```
